Probabilistic Catalogs from Crowded Star Fields

### 2012-09-11 paper/ms.tex
Broken power-law plot
Writing

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single single answer to this question (i.e. a single catalog), but rather a
probability distribution over the space of possible catalogs. This is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
catalogs.

Bayesian object detection has been implemented by \citet{2011MNRAS.415.3462F}
under the assumption of a known $N$. This assumption is required for the
{\tt MultiNest} sampler to be applicable. Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}
We also model how the parameters give rise to the data, via a {\it sampling
distribution}:
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions, one for each possible value of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

When specific data $x^*$ has been obtained, our state of knowledge about $\theta$
gets updated from the prior to the posterior,
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The modulating function is the {\it likelihood function}, which is the prior
probability of obtaining the actual data set $x^*$ as a function of the
parameters\footnote{In the case that the sampling distribution is a probability
density, the likelihood is the probability density at the observed data. This
is usually fine, although one should be aware of the Borel-Kolmogorov ``paradox''}.
As suggested by the above notation, it is obtained from the
sampling distribution with the actual data substituted in.

@@

\subsection{The Hypothesis Space and the Prior}
Our model makes the following assumptions. There are an unknown number of stars
$N$ in the field. Each star has an unknown
position $(x,y)$ in the plane of the sky, and an unknown flux $f$. We also
describe the {\it spatial distribution} of the stars by parameters
$\alpha$, and the distribution of fluxes (commonly known as the {\it luminosity
function}) by parameters $\beta$. In summary, the unknown parameters are:
\begin{eqnarray}
\theta = \left\{N, \alpha, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}
\end{eqnarray}
The prior probability distribution for the unknown parameters can be factorized
using the product rule. With a variety of independence assumptions, the prior
can be written as:
\begin{eqnarray}
p(\theta) = p(N)p(\alpha)p(\beta)\prod_{i=1}^N p(x_i, y_i | \alpha)
p(f_i | \beta) 
\end{eqnarray}
Here, we have assumed that the luminosity function does not depend on position.
We have also assumed independence of $N$, $\alpha$ and $\beta$, which means that
learning the value of one would not tell us anything about the values of the
others. Finally, the positions and fluxes of the stars come independently from
a common distribution. If we knew the spatial distribution and the luminosity
function of the stars, then the location and flux of a particular star would
not tell us anything about the location and flux of another star. Really, this
is just a way of implementing exchangeability.

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data. The pixel-convolved PSF-convolved scene is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved PSF. Throughout this paper we
assume the pixel-convolved PSF is a known mixture of two concentric circular
Gaussians:
\begin{eqnarray}
\mathcal{P}(x, y) &=&
\end{eqnarray}

\subsection{The Prior Distribution}
A model for the spatial distribution corresponds to a choice of functional
forms for the distributions $p(x_i, y_i | \alpha)$ and $p(f_i | \beta)$.

\section{Broken Power-Law Distribution}
The broken power-law distribution is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
x^{-\alpha_1 - 1} & & x_1 \leq x \leq x_2 \\
x^{-\alpha_2 - 1} & & x > x_2
\end{array}
\right.
.
\end{eqnarray}
With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & x_1 \leq x \leq x_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > x_2
\end{array}
\right.
.
\end{eqnarray}

@@

>>>>

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single catalog), but rather a
probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should be in the catalog) is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
catalogs.

Bayesian object detection has been implemented by \citet{2011MNRAS.415.3462F}
under the assumption of a known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}.
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the data space, one for each possible value
of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

When specific data $x^*$ has been obtained, our state of knowledge about $\theta$
gets updated from the prior distribution to the posterior distribution,
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The term $p(\theta)p(x|\theta)|_{x=x^*} = \mathcal{L}(\theta; x)$
is the {\it likelihood function}, which is the 
probability of obtaining the actual data set $x^*$ as a function of the
parameters\footnote{In the case that the sampling distribution is a probability
density, the likelihood is the probability density evaluated at the observed
data. This usually causes no problems, although one should be aware of the
Borel-Kolmogorov paradox \citep{jaynes}.}.
As suggested by the above notation, the likelihood function is obtained from the
sampling distribution with the actual data substituted in.

@@

\subsection{The Hypothesis Space and the Prior}
Our model makes the following assumptions. There are an unknown number of stars
$N$ in the field. Each star has an unknown
position $(x,y)$ in the plane of the sky, and an unknown flux $f$. We also
describe distribution of fluxes (commonly known as the {\it luminosity
function}) of the stars by parameters $\beta$. In summary, the unknown
parameters are:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}
\end{eqnarray}
The prior probability distribution for the unknown parameters can be factorized
using the product rule. With a variety of independence assumptions, the prior
can be written as:
\begin{eqnarray}
p(\theta) = p(N)p(\beta)\prod_{i=1}^N p(x_i, y_i)
p(f_i | \beta) 
\end{eqnarray}
Here, we have assumed that the luminosity function does not depend on position.
We have also assumed independence of $N$ and $\beta$, which means that
learning the value of one would not tell us anything about the values of the
others. Finally, the fluxes of the stars come independently from
a common distribution. If we knew the luminosity
function of the stars, then the location and flux of a particular star would
not tell us anything about the location and flux of another star. Really, this
is just a way of implementing exchangeability of the stars.

For simplicity, we assume a uniform prior probability distribution for the
position of each star. This creates a strong preference for catalogs where
the stars are uniformly distributed across the image. Thus, this model is appropriate for small patches of sky where the density of stars is approximately
uniform. In other scenarios, such as images of stellar clusters, it is possible
to parameterise the spatial distribution of the stars in a similar way to how
we have parameterised the luminosity function.

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. Thus, we need
a prescription for simulating images from a catalog:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}
\end{eqnarray}
If we knew the true catalog, we could compute the image we would expect to see,
in the absence of noise. This mock image is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we
assume the pixel-convolved PSF is a known weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The observed image is assumed to be generated from the mock image (evaluated
at the center of each pixel) plus Gaussian noise:
\begin{eqnarray}
y_{ij} &=& \mathcal{M}(x_{ij}, y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where $\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$.

\subsection{The Prior Distribution}
The prior distribution for the number of stars $N$ is uniform between 0 and
$N_{\rm max}$.

\section{Broken Power-Law Distribution}
The broken power-law distribution is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
x^{-\alpha_1 - 1} & & x_1 \leq x \leq x_2 \\
x^{-\alpha_2 - 1} & & x > x_2
\end{array}
\right.
.
\end{eqnarray}

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{Figures/broken.eps}
\end{center}
\end{figure}

With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & x_1 \leq x \leq x_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > x_2
\end{array}
\right.
.
\end{eqnarray}

@@

\bibitem[\protect\citeauthoryear{Feroz, Hobson, 
\& Bridges}{2009}]{multinest} Feroz F., Hobson M.~P., Bridges M., 2009, MNRAS, 398, 1601
